Modal Abstractions for Virtualizing Memory Addresses

### 2023-04-05 examples.tex
+

\section{Experiment}
\label{sec:experiment}
\mytodo{Mechanization is Ismail's Part -- Discussing and putting in here is Colin's}

@@

\subsection{Identity Mappings}
\todo[inline]{Identity mappings are difficult, and our current approach won't quite work. Consider trying to have a virtual pointsto for an actual page table entry (i.e., that one could use to update a page table mapping), while also having a virtual pointsto for an address that entry mapped. With the current (let's call it v1) solution, we can't actually have both of those simultaneously!  That's because the PTE pointsto will assert full ownership of the physical memory cell holding the PTE as its data value, while the virtual pointsto for the data mapped by that entry will \emph{also} assert (fractional) ownership of all entries a page table walk would traverse.
}
\todo[inline,color=violet]{This doesn't seem to cause issues with the mapping/unmapping examples, only with changing intermediate page table pointers. The mapping example requires a virtual pointsto for the blank PTE, and once filled in that ownership can be immediately split to create the 512 new virtual pointsto assertions for the newly mapped page. Conversely, for unmapping we'd assume ownership of all the relevant virtual pointsto assertions for the page we're unmapping, at which point we can (with a bit of work) show that they all correspond to the same L1 PTE, and extract the 512 fractional shares of that entry from the pointsto assertions.  But changing intermediate page tables, as one would do for coallescing or splitting a superpage while preserving the virtual-to-physical mappings, couldn't be done without some really complicated separating implication tricks.}
\todo[inline,color=green]{One possible approach to resolving this, which we came up with in our Tuesday meeting, is to recognize that the current (v1) virtual points-to is too strong, because it really doesn't care about \emph{owning} those fractional resources, it only cares that \emph{something} ensures the correct page table walk exists. Iris has a ghost map resource where authoritative ownership of an individual key-value pair can be handled as a resource.  (Colin was using this in the filesystem cache.)
We can use that mechanism to separate the virtual-to-physical translation from the physical memory involved (Kolanski and Klein may have done something similar for different reasons): (fractional) virtual points-to assertions can be defined in terms of (fractional) ownership of these authoritative ghost map entry assertions, plus sharing an invariant that the current installed page table respects all entries of the mapping. Unmapping collects the authoritative map kvpairs from collecting the assertions, and then can remove them from the ghost map and update the page tables. Critically, physical ownership of the page tables then lives in the invariant on the current page table, so some virtual pointsto assertions can refer to memory in those page tables.
This still works with the modality, since that invariant is also semantically a predicate on a page table root.
Let's call this v2.
}
\todo[inline]{the current course of action is to finish the WPs for v1 and try an example that should work with either approach (like the task switch). This is a good basic sanity check on our approaches. Once that's done, then we work out the details of v2, which should require \emph{refactoring} the same sorts of reasoning we had for v1, rather than totally different proofs from v1.}

\subsection{Mapping a New Page}
One of the key tasks of a page fault handler in a general-purpose OS kernel is to map new pages into an address space by writing into an existing page table.
\todo[inline]{Ismail, see notes about intermediate assertions on the Notion page}
\begin{figure}
\begin{lstlisting}
;; Assume the faulting address f is in r12,
;; a callee-save register, and f with the 
;; lower 12 bits zeroed out (so, the virtual
;; page containing f) is in r13 (also callee-save)

;; Ensure there's a place to put the L1 mapping
mov %rdi, %r13
call ensure_L1_page
;; Returns the virtual address of the L1 entry in rax
mov %r14, %rax ;; Save that before another call
call alloc_phys_page_or_panic
;; Calculate new PTE by setting the low-order bits
;; of the physical page
add %rax, 3
;; update the page table entry, mapping the page
mov (%r14), %rax
;; memset the newly-mapped page
mov %rdi, %r13
mov %rsi, 0
mov %rdx, 4096
call memset
;; The newly mapped page is full of 0s
\end{lstlisting}
\caption{Distilled code for mapping a new page}
\label{fig:mapping_code}
\todo[inline]{this is terribly ugly, should switch to minted}
\end{figure}

>>>>

%$\assert{\ulcorner \mathsf{aligned maddr} \urcorner \ast \mathsf{r14} \mapsto_{\textsf{r}} \textsf{r14v} \ast \mathsf{r13} \mapsto_{\textsf{r}} \textsf{maddr} \ast \mathsf{rdi} \mapsto_{\textsf{r}} \textsf{rdiv} \ast \mathsf{rax} \mapsto_{\textsf{r}} \textsf{raxv}}$
%$\assert{\ulcorner \ptablestore !! \textsf{maddr} = \textsf{None} \urcorner \ast \ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}} \ast \textsf{Pf}} $
\section{Experiment}
\label{sec:experiment}
\mytodo{Mechanization is Ismail's Part -- Discussing and putting in here is Colin's}

@@

\begin{comment}
\todo[inline]{Identity mappings are difficult, and our current approach won't quite work. Consider trying to have a virtual pointsto for an actual page table entry (i.e., that one could use to update a page table mapping), while also having a virtual pointsto for an address that entry mapped. With the current (let's call it v1) solution, we can't actually have both of those simultaneously!  That's because the PTE pointsto will assert full ownership of the physical memory cell holding the PTE as its data value, while the virtual pointsto for the data mapped by that entry will \emph{also} assert (fractional) ownership of all entries a page table walk would traverse.
}
\todo[inline,color=violet]{This doesn't seem to cause issues with the mapping/unmapping examples, only with changing intermediate page table pointers. The mapping example requires a virtual pointsto for the blank PTE, and once filled in that ownership can be immediately split to create the 512 new virtual pointsto assertions for the newly mapped page. Conversely, for unmapping we'd assume ownership of all the relevant virtual pointsto assertions for the page we're unmapping, at which point we can (with a bit of work) show that they all correspond to the same L1 PTE, and extract the 512 fractional shares of that entry from the pointsto assertions.  But changing intermediate page tables, as one would do for coallescing or splitting a superpage while preserving the virtual-to-physical mappings, couldn't be done without some really complicated separating implication tricks.}
\todo[inline,color=green]{One possible approach to resolving this, which we came up with in our Tuesday meeting, is to recognize that the current (v1) virtual points-to is too strong, because it really doesn't care about \emph{owning} those fractional resources, it only cares that \emph{something} ensures the correct page table walk exists. Iris has a ghost map resource where authoritative ownership of an individual key-value pair can be handled as a resource.  (Colin was using this in the filesystem cache.)
We can use that mechanism to separate the virtual-to-physical translation from the physical memory involved (Kolanski and Klein may have done something similar for different reasons): (fractional) virtual points-to assertions can be defined in terms of (fractional) ownership of these authoritative ghost map entry assertions, plus sharing an invariant that the current installed page table respects all entries of the mapping. Unmapping collects the authoritative map kvpairs from collecting the assertions, and then can remove them from the ghost map and update the page tables. Critically, physical ownership of the page tables then lives in the invariant on the current page table, so some virtual pointsto assertions can refer to memory in those page tables.
This still works with the modality, since that invariant is also semantically a predicate on a page table root.
Let's call this v2.
}
\todo[inline]{the current course of action is to finish the WPs for v1 and try an example that should work with either approach (like the task switch). This is a good basic sanity check on our approaches. Once that's done, then we work out the details of v2, which should require \emph{refactoring} the same sorts of reasoning we had for v1, rather than totally different proofs from v1.}
\end{comment}
\subsection{Mapping a New Page}
One of the key tasks of a page fault handler in a general-purpose OS kernel is to map new pages into an address space by writing into an existing page table.
\todo[inline]{Ismail, see notes about intermediate assertions on the Notion page}
\begin{figure}
\begin{lstlisting}
;; Assume the faulting address f is in r12,
;; a callee-save register, and f with the 
;; lower 12 bits zeroed out (so, the virtual
;; page containing f) is in r13 (also callee-save)

;; Ensure there's a place to put the L1 mapping
mov %rdi, %r13
call ensure_L1_page
;; Returns the virtual address of the L1 entry in rax
mov %r14, %rax ;; Save that before another call
call alloc_phys_page_or_panic
;; Calculate new PTE by setting the low-order bits
;; of the physical page
add %rax, 3
;; update the page table entry, mapping the page
mov (%r14), %rax
\end{lstlisting}
\caption{Distilled code for mapping a new page}
\label{fig:mapping_code}
\todo[inline]{this is terribly ugly, should switch to minted}
\end{figure}
